---
title: "Importing Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(honupsy)
```

## Overview

The honupsy package imports, validates and summarises data from the Swiss
national quality measurement in adult inpatient psychiatry, coordinated by the
ANQ (Nationaler Verein für Qualitätsentwicklung in Spitälern und Kliniken).

This vignette covers the import step. All other analyses in the package build
on the data structures produced here, so it is worth reading even if the
technical details of file formats are not your main interest.

## Works directly with submission files

Swiss psychiatric clinics are required by law to collect and submit these data
annually. The data are submitted in a defined format -- either as text files
or using the official ANQ data-entry template (an Excel file provided by ANQ).

**honupsy works directly with these submission files.** There is no need to
reformat or restructure your data before importing. The same files you send
to ANQ can be loaded straight into R.

Importing needs nothing further, but the analysis does: the `summarise_*()`
functions operate per unit (ward), and the ANQ format carries no unit
identifier. Unit assignment is therefore a required preparation step before any
summary -- you supply it separately, see
[Adding unit assignments](#adding-unit-assignments) below.

## What data are imported

The ANQ submission for adult inpatient psychiatry consists of five record
types, spanning two layers: the administrative case data drawn from the BFS
Medical Statistics (Medizinische Statistik der Krankenhäuser, collected by the
Bundesamt für Statistik, BFS), and the ANQ measurement instruments collected
on top of it.

| Record type | Content | Layer |
|---|---|---|
| MB | Minimal dataset (administrative and clinical case data) | BFS Medical Statistics |
| MP | Additional psychiatric case data | BFS Medical Statistics |
| PH | HoNOS symptom burden at admission and discharge | ANQ measurement |
| PB | BSCL patient self-report at admission and discharge | ANQ measurement |
| FM | Coercive measures during inpatient stay | ANQ measurement |

The record types are linked by the case identifier (FID), present in every row.
`import_anq()` imports all five and returns them as a named list of data
frames, one per record type.

## Supported file formats

honupsy supports three file formats:

**Text files** (`.txt`) with pipe (`|`), semicolon (`;`) or tab as delimiter --
the delimited export format of the ANQ submission.

**Simple XLSX** with one worksheet per record type, named MB, MP, PH, PB
and FM.

**The official ANQ data-entry template** -- the Excel file provided by ANQ
for manual data entry. honupsy automatically recognises this file by its
worksheet names and handles the multi-row header structure.

## Basic usage

```{r, eval = FALSE}
library(honupsy)

# import from a single mixed text file
data <- import_anq("path/to/data.txt")

# import from the official ANQ data-entry template
data <- import_anq("path/to/ANQ_PSY_EP_Dateneingabe_Messjahr-2024.xlsx")

# import from separate files per record type
data <- import_anq(c(
  MB = "path/to/mb.txt",
  MP = "path/to/mp.txt",
  PH = "path/to/ph.txt",
  PB = "path/to/pb.txt",
  FM = "path/to/fm.txt"
))
```

It is also possible to import only a subset of record types:

```{r, eval = FALSE}
# import only MB and FM
data <- import_anq(c(
  MB = "path/to/mb.txt",
  FM = "path/to/fm.txt"
))
```

Record types not present in the input are returned as empty data frames with
the correct column types, so downstream code does not need to check for
missing datasets.

## The import summary

After a successful import, honupsy prints a brief summary:

```{r}
path <- system.file("extdata", "example.txt", package = "honupsy")
data <- import_anq(path, validate = FALSE)
```

This tells you at a glance how many cases, assessments and measures were
found in the data, and how many assessments were recorded as dropouts.

## The returned data structure

`import_anq()` returns a named list with five elements:

```{r}
names(data)
```

Each element is a data frame:

```{r}
# one row per case
nrow(data$mb)

# two rows per case: admission and discharge
nrow(data$ph)

# zero to many rows per case
nrow(data$fm)
```

The data frames can be used directly with standard R tools and are compatible
with tidyverse packages.

A few selected columns to illustrate the structure:

```{r}
# MB: BFS case data (administrative and clinical)
data$mb[, c("fid", "sex", "age_admission", "main_diagnosis",
            "admission", "discharge")]
```

```{r}
# PH: ANQ measurement (HoNOS assessments)
data$ph[, c("fid", "time_point_label", "is_dropout",
            "honos_total", "honos_n_valid")]
```

```{r}
# FM: ANQ measurement (coercive measures)
data$fm[, c("fid", "measure_label", "start_date",
            "duration_min", "is_coercive_medication")]
```

## Data validation

By default, `import_anq()` runs a plausibility check on the imported data
and issues warnings if problems are found. The check covers:

- Completeness: are all expected record types present?
- FID consistency: do the case identifiers match across record types?
- Content plausibility: are values in the expected ranges?

```{r, eval = FALSE}
# validation is on by default
data <- import_anq("path/to/data.txt")

# inspect the validation messages programmatically
data$.validation
```

If you are confident in your data and want to suppress the validation step,
use `validate = FALSE`:

```{r, eval = FALSE}
data <- import_anq("path/to/data.txt", validate = FALSE)
```

## A note on PB (BSCL)

The BSCL patient self-report (PB) is not collected by all clinics. Forensic and
old-age psychiatry units are exempt from this measurement. If your data do not
contain PB records, `data$pb` will be an empty data frame -- this is expected
and does not trigger a validation warning.

## Adding unit assignments

The ANQ submission format does not include a unit or ward identifier. If your
analyses require unit-level breakdowns, you can supply a separate lookup table
that maps each case to a unit.

The lookup table can be an Excel file, a CSV, or a delimited text file. The
column names do not need to follow any particular convention – you specify
which columns contain the FID and the unit identifier:

```{r units}
# load unit assignments from the example Excel file
# column names reflect a typical clinic export
units_path <- system.file("extdata", "example_units.xlsx",
                           package = "honupsy")

units <- import_unit_assignments(
  units_path,
  col_fid  = "Fallnummer",
  col_unit = "Station"
)

units
```

Attach the unit assignments to the imported data:

```{r assign-units}
data <- assign_units(data, units)
```

A `unit` column is now available in all five datasets:

```{r check-units}
data$mb[, c("fid", "unit", "main_diagnosis")]
data$ph[, c("fid", "unit", "time_point_label", "honos_total")]
```

If a patient transferred between units during their stay, we recommend using
the unit the patient was on immediately before discharge.

## Scope and versions

honupsy covers adult inpatient psychiatry (Erwachsenenpsychiatrie). The child
and adolescent instruments (HoNOSCA, HoNOSCA-SR) are out of scope.

The data definition has changed across measurement years. From measurement
year 2026 the case definition adopts case merging, and SpiGes replaces the
former Medical Statistics. If you work with recent data, check your files
against the data definition for the relevant measurement year.

## Next steps

Once the data are imported and unit assignments are attached, the data frames
can be used directly for analysis in R. Unit-level descriptive summaries are 
available through `summarise_composition()`, `summarise_honos_severity()`, and 
`summarise_occupancy()`.
