---
title: "Importing ANQ Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing ANQ Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(honupsy)
```

## Overview

The honupsy package provides tools to import, validate and analyse data from
the Swiss national quality measurement programme in inpatient psychiatry
(Nationale Qualitätsmessung stationäre Psychiatrie, ANQ).

This vignette covers the import step. All other analyses in the package build
on the data structures produced here, so it is worth reading even if the
technical details of file formats are not your main interest.

## No additional data preparation required

Swiss psychiatric clinics are required by law to collect and submit ANQ data
annually. The data are submitted in a defined format -- either as text files
or using the official ANQ data entry template (an Excel file provided by ANQ).

**honupsy works directly with these submission files.** There is no need to
reformat or restructure your data before importing. The same files you send
to ANQ can be loaded straight into R.

## What data are imported

The ANQ measurement for adult inpatient psychiatry consists of five record
types:

| Record type | Content |
|---|---|
| MB | Minimal dataset (administrative and clinical data per case) |
| MP | Additional psychiatric data per case |
| PH | HoNOS symptom burden assessment at admission and discharge |
| PB | BSCL patient self-report at admission and discharge |
| FM | Coercive measures during inpatient stay |

`import_anq()` imports all five record types and returns them as a named list
of data frames, one per record type.

## Supported file formats

honupsy supports three file formats:

**Text files** (`.txt`) with pipe (`|`), semicolon (`;`) or tab as delimiter --
the format used for the mandatory ANQ submission to the Federal Statistical
Office (BFS).

**Simple XLSX** with one worksheet per record type, named MB, MP, PH, PB
and FM.

**The official ANQ data entry template** -- the Excel file provided by ANQ
for manual data entry. honupsy automatically recognises this file by its
worksheet names and handles the multi-row header structure.

## Basic usage

```{r, eval = FALSE}
library(honupsy)

# import from a single mixed text file
data <- import_anq("path/to/data.txt")

# import from the official ANQ data entry template
data <- import_anq("path/to/ANQ_PSY_EP_Dateneingabe_Messjahr-2024.xlsx")

# import from separate files per record type
data <- import_anq(c(
  MB = "path/to/mb.txt",
  MP = "path/to/mp.txt",
  PH = "path/to/ph.txt",
  PB = "path/to/pb.txt",
  FM = "path/to/fm.txt"
))
```

It is also possible to import only a subset of record types:

```{r, eval = FALSE}
# import only MB and FM
data <- import_anq(c(
  MB = "path/to/mb.txt",
  FM = "path/to/fm.txt"
))
```

Record types not present in the input are returned as empty data frames with
the correct column types, so downstream code does not need to check for
missing datasets.

## The import summary

After a successful import, honupsy prints a brief summary:

```{r}
path <- system.file("extdata", "example.txt", package = "honupsy")
data <- import_anq(path, validate = FALSE)
```

This tells you at a glance how many cases, assessments and measures were
found in the data, and how many assessments were recorded as dropouts.

## The returned data structure

`import_anq()` returns a named list with five elements:

```{r}
names(data)
```

Each element is a data frame:

```{r}
# one row per case
nrow(data$mb)

# two rows per case: admission and discharge
nrow(data$ph)

# zero to many rows per case
nrow(data$fm)
```

The data frames can be used directly with standard R tools and are compatible
with tidyverse packages.

A few selected columns to illustrate the structure:

```{r}
# MB: administrative and clinical data
data$mb[, c("fid", "sex", "age_admission", "main_diagnosis",
            "admission", "discharge")]
```

```{r}
# PH: HoNOS assessments
data$ph[, c("fid", "time_point_label", "is_dropout",
            "honos_total", "honos_n_valid")]
```

```{r}
# FM: coercive measures
data$fm[, c("fid", "measure_label", "start_date",
            "duration_min", "is_coercive_medication")]
```

## Data validation

By default, `import_anq()` runs a plausibility check on the imported data
and issues warnings if problems are found. The check covers:

- Completeness: are all expected record types present?
- FID consistency: do the case identifiers match across record types?
- Content plausibility: are values in the expected ranges?

```{r, eval = FALSE}
# validation is on by default
data <- import_anq("path/to/data.txt")

# inspect the validation messages programmatically
data$.validation
```

If you are confident in your data and want to suppress the validation step,
use `validate = FALSE`:

```{r, eval = FALSE}
data <- import_anq("path/to/data.txt", validate = FALSE)
```

## A note on PB (BSCL)

The BSCL patient self-report (PB) is not collected by all clinics. Forensic
psychiatric institutions have been exempt from this measurement since
1 July 2019. If your data do not contain PB records, `data$pb` will be an
empty data frame -- this is expected and does not trigger a validation
warning.

## Next steps

Once your data are imported you can proceed to descriptive summaries,
change score analyses between admission and discharge, and the estimation
procedures provided by honupsy. These are covered in the other vignettes.
