---
title: "Summarising Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summarising Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(honupsy)
```

## Overview

The *Importing Data* vignette covers getting submission files into R. This
one picks up from there and shows the descriptive `summarise_*()` functions:
case composition, HoNOS symptom severity at admission, and unit occupancy.

To keep the examples free of real patient data, we generate a synthetic cohort
with `sample_cases()` and run the summaries over it. The functions behave
exactly as they would on imported clinic data.

## Why synthetic data

`sample_cases()` does not fake the import structure. It writes valid ANQ raw
records to a temporary file and reads them back through the same parser as
`import_anq()`, so the result cannot drift from the real import format. That
makes it suitable for examples, teaching, and tests without touching protected
data.

```{r generate}
set.seed(1)

data <- sample_cases(n = 1000)

names(data)
nrow(data$mb)
```

`sample_cases()` currently generates only MB (case-level data) and PH (HoNOS
assessments).

> **Note (MP).** `mp` is returned as an empty data frame. No summariser uses
> the additional psychiatric data yet; this will change later.

> **Note (FM).** `fm` is also empty, so there is no coercive-measures summary
> in this vignette yet. Planned for a later version.

## Attaching unit assignments

The ANQ format carries no ward identifier, and the `summarise_*()` functions
work per unit, so unit assignments must be attached first. Here we build a
lookup table in memory; in practice it comes from a clinic export (see
`import_unit_assignments()` in the import vignette).

```{r assign-units}
set.seed(1)

lookup <- data.frame(
  fid  = data$mb$fid,
  unit = sample(c("Ward A", "Ward B", "Ward C"), nrow(data$mb), replace = TRUE)
)

data <- assign_units(data, import_unit_assignments(lookup))

data$mb[1:5, c("fid", "unit", "main_diagnosis")]
```

## Case composition

`summarise_composition()` summarises the BFS case layer (MB). It returns one row
per unit with case counts, mean age, mean length of stay, and the female
proportion. The `n_*_missing` columns report how many cases were excluded from
each statistic for missingness.

```{r composition}
summarise_composition(data)
```

`detail = "full"` adds the full distribution (median, SD, min, max, quartiles,
IQR) for age and length of stay.

```{r composition-full}
summarise_composition(data, detail = "full")
```

## HoNOS severity at admission

`summarise_honos_severity()` summarises the ANQ measurement layer (PH):
admission HoNOS assessments (time point 1, non-dropout) per unit -- mean score
per item, the proportion of severe cases per item (score > 2), and the mean
total score.

The leading columns are the data-quality guardrails. `prop_assessed` is the
share of cases with a usable admission assessment, and `mean_items_rated` is the
average number of the 12 items actually scored. Read the severity figures in
light of these: a low `prop_assessed` means the means describe only part of the
unit.

```{r honos}
honos <- summarise_honos_severity(data)

# quality block plus the total score
honos[, c("unit", "n_cases", "n_honos_adm", "prop_assessed",
          "mean_items_rated", "n_partial", "honos_total_mean")]
```

The per-item means and severe proportions are in the remaining columns:

```{r honos-items}
honos[, c("unit", "h1_mean", "h2_mean", "h6_mean",
          "h1_prop_severe", "h2_prop_severe", "h6_prop_severe")]
```

## Unit occupancy

`summarise_occupancy()` gives annual per-unit occupancy: estimated unit size
(1% rule), mean and maximum daily census, and the corresponding occupancy rates.
It also reports how many cases were dropped, and why.

```{r occupancy}
occ <- summarise_occupancy(data, year = 2023)

occ$by_unit

occ$excluded
```

For the underlying daily series -- for example to plot occupancy over the year
-- call `occupancy_daily()` directly. It returns the per-day census per unit.

```{r occupancy-daily}
daily <- occupancy_daily(data, year = 2023)

head(daily$daily)
daily$unit_size
```

## Exercising the data-quality columns

On clean data the missingness and dropout columns stay near zero. Their purpose
is visible only when the data are imperfect, which is the realistic case. The
probability arguments to `sample_cases()` let you dial in dropouts, unrated
items, and missing discharge dates.

```{r messy}
set.seed(2)

messy <- sample_cases(
  n = 1000,
  p_dropout_adm       = 0.10,  # 10% of admission HoNOS are dropouts
  p_item_na           = 0.05,  # 5% of rated items coded 9 (not known)
  p_missing_discharge = 0.08   # 8% of cases lack a discharge date
)

messy_lookup <- data.frame(
  fid  = messy$mb$fid,
  unit = sample(c("Ward A", "Ward B", "Ward C"), nrow(messy$mb), replace = TRUE)
)
messy <- assign_units(messy, import_unit_assignments(messy_lookup))
```

The dropouts and unrated items now move `prop_assessed`, `mean_items_rated`, and
`n_partial`:

```{r messy-honos}
summarise_honos_severity(messy)[, c("unit", "n_cases", "n_honos_adm",
                                     "prop_assessed", "mean_items_rated",
                                     "n_partial")]
```

The missing discharge dates show up in the occupancy exclusion tally:

```{r messy-occupancy}
summarise_occupancy(messy, year = 2023)$excluded
```

## Next steps

These summaries operate on MB and PH. Summaries drawing on the additional
psychiatric data (MP), the BSCL self-report (PB), and coercive measures (FM) are
planned for later versions, at which point `sample_cases()` will generate those
record types as well.
